---
title: "RaukR project"
subtitle: "Analysis of genomic array data from 9 dog breeds"
author: Cátia Pereira, Dovilė Sinkevičiūtė, Minna Sivonen and Svandís Sæmundsdóttir
date: "`r format(Sys.Date(),format='%d/%m/%Y')`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    theme: flatly
    highlight: kate
    df_print: paged
    code_folding: hide
    self_contained: false
    keep_md: false
    encoding: "UTF-8"
    css: ["assets/lab.css"]
---

```{r include=FALSE}
## Load packages to use for the data analysis
library(tidyr)
library(dplyr)
library(ggplot2)
library(GenABEL)

# Load the data
load("geno_data_raw.rdat")
```


# Description of the dataset

## Nine dog breeds from five different European countries were analysed in this study.{.tabset}
```{r include=FALSE}
length(unique(data_raw@phdata$Country)) ## data from 5 countries
length(unique(data_raw@phdata$Breed)) ## data covers 9 dog breeds
```
### Dog breeds in the study
```{r}
# number of dogs per breed (male, female)
dogs <- data_raw@phdata %>%
  group_by(Breed, sex) %>%
  summarize(n = n()) %>%
  group_by(Breed) %>%
  spread(key="sex", value = "n" ) %>%
  mutate_all(~replace(., is.na(.), 0))

dogs$Breed_name <- c("Boxer",
                     "Belgian Shepperd",
                     "Cavalier King Charles spaniel",
                     "Dachshund", 
                     "Doberman pinscher",
                     "Finnish lapphund",
                     "German Shepperd",
                     "Labrador retriever", 
                     "Newfoundland")

colnames(dogs)<- c("Abbreviation", "Female", "Male", "Breed")
print(dogs[,c("Breed", "Abbreviation", "Female", "Male")])

```
### Countries
```{r}
country_info <- data_raw@phdata %>% 
  group_by(Country) %>%
  summarize(total_dogs = sum(n())) %>%
  arrange(desc(total_dogs))

country_info$fullname <- c("Finland", "Belgium", "Sweden", "France", "Denmark") 
colnames(country_info)<- c("Abbreviation", "Total number of dogs", "Country")
print(country_info[,c("Country", "Abbreviation", "Total number of dogs")])
```

### Distribution of breeds by country
```{r}
data_raw@phdata %>%
  group_by(Breed, Country) %>%
  summarize(Dogs = n()) %>%
  arrange(desc(Dogs)) %>%
  arrange(desc(Country))
```

## Weight {.tabset}
### Table
Table with the mean age and weight of the dogs, taking sex into account
```{r}
# mean ans sd functions
funm <- function(x){mean(x, na.rm = T)}

#table
dogsWH <- data_raw@phdata %>%
  group_by(Breed, sex) %>%
  summarize(n = n(),
            mean.Age = funm(Age), 
            mean.Weight = funm(Body_weight)) 

dogsWH$sex[dogsWH$sex == 0]<- "female"
dogsWH$sex[dogsWH$sex == 1]<- "male"

print(dogsWH)
```

### Graph
Weight distribution for all breeds
```{r}
data_raw@phdata %>% 
    dplyr::select(Breed, Body_weight) %>% 
    na.omit() %>% 
    ggplot(aes(x=Breed, y=Body_weight, fill=Breed)) +
    geom_boxplot() +
    geom_jitter(width = 0.2, alpha=0.3) +
    labs(title="Weight distribution for all breeds",
         subtitle="RaukR project 2019",
         caption="Sources: PLoS ONE, 2015")+
    theme_bw() + 
    theme(legend.position = "none")
```


#Principal Component Analysis
```{r, echo=FALSE}
# Find marker for PCA analysis
# qc0 <- check.marker(data_raw, call = 0.95, perid.call=0.95,
#                     maf=1e-06, p.lev=1e-08, ibs.exclude="lower")
# # Save output from check.marker
# data.qc0 <- data_raw[qc0$idok, qc0$snpok]
# 
# # Remove chromosome 39 (X) from dataset
# autosomalMarkerNames <- snpnames(data.qc0)[chromosome(data.qc0) != 39]
# # Compute genomic kinship matrix
# data.qc0.gkin <- ibs(data.qc0[, autosomalMarkerNames], weight = "freq")
# # Transform it to a distance matrix
# data.qc0.dist <- as.dist(0.5 - data.qc0.gkin)
# # Perform multidimensional scaling to display individuals on a 2D
# # plot preserving genomic distances between them
# data.qc0.mds <- cmdscale(data.qc0.dist)
# # Save the results so no need to run every single time
# pca_result <- save(data.qc0.mds, file="pca_results.rdat")

# Plot the result
load("pca_results.rdat")
df_data_raw <- data.frame(Dim_1 = data.qc0.mds[,1], Dim_2 = data.qc0.mds[,2], Breed= data_raw@phdata$Breed)
ggplot(df_data_raw, aes(Dim_1, Dim_2, color=Breed)) + 
  geom_point(alpha = 0.8) + 
  labs(title="Principal Component Analysis",
       subtitle="RaukR project 2019") +
  theme_bw() +
  scale_color_manual(values = c("black", "dodgerblue4", "bisque4", "darkgreen","yellow3", "darkgoldenrod3", "darkorchid4", "darkred", "grey")) 
```

#Multidimensional scaling MDS
```{r, cache= TRUE, include=FALSE}
# Compute genomic kinship-based distances
gkin <- ibs(data_raw, weight = 'freq')
head(gkin)
dm <- as.dist(.5 - gkin) # Normalize it

# PCA
pca <- prcomp(dm) #Principal components analysis

library(devtools)
install_github("vqv/ggbiplot") #to get the ggbiplot
library(ggbiplot)

#vector for the colors
color_9 <- c("black", "dodgerblue4", "bisque4", "darkgreen","yellow3", "darkgoldenrod3", "darkorchid4", "darkred", "grey")

g <- ggbiplot(pca, obs.scale = 1, var.scale = 1,
              groups = data_raw@phdata$Breed, ellipse = F, #using breed as a group
              circle = TRUE, var.axes = F) +
  ggtitle("PCA")+
  scale_color_manual(values = color_9)+
  theme(legend.position = 'none') #removing legend

# MDS
ibs <- as.data.frame(cmdscale(dm))

ibs <- cbind(ibs, pop = data_raw@phdata$Breed) #same as group at PCA

z <-ggplot(ibs, mapping = aes(x=V1, y=V2, col=pop)) + 
  geom_point() +
  ggtitle("MDS") +
  scale_color_manual(values = color_9) +
  theme(legend.position = 'none') #removing legend

library(gridExtra)
library(grid)
library(ggplot2)
library(lattice)
library(ggpubr)

#arranging the two grids side by side with one legend
ggarrange(g,z, ncol=2, nrow=1, common.legend = TRUE, legend="bottom")

```

